---
title: "Marginal Structural Models: A MIMIC-III Workshop"
author: "C.V. Cosgriff, MD, MPH"
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

For this example we'll use the dataset which was used in the following
study by Hsu and collegues:

_Hsu DJ, Feng M, Kothari R, Zhou H, Chen KP, Celi LA. The association between indwelling arterial catheters and mortality in hemodynamically stable patients with respiratory failure: a propensity score analysis. CHEST Journal. 2015 Dec 1;148(6):1470-6._

We previously extracted the requisite data, and will load it now along with the
requisite libraries.
```{r message=FALSE, error=FALSE}
library(dplyr)
library(readr)
library(sandwich)

aline_tbl <- read_csv('./aline_data.csv')
```

We will be focusing on the following research question: __what is the _causal effect_
of indwelling cathethers and mortality in hemodynamically stable patients with
respiratory failure?__

The exposure, $A$, will be `aline_flg`, a binary variable indicating whether or
not the patient had an indwelling arterial cathether. 

$L$ will be a vector composed of the variables age, gender, weight on first day, 
SOFA score, whether the patient was on a surgical service, time in the ICU, 
indicators for the presence of disease including CHF, AFib, CKD, liver disease,
COPD, CAD, stroke, malignancy, and first day vital sign and laboratory data.  

We will now format our data. We first designate our categorical variables, and
then we exclude all unnecessary variables from our dataset.
```{r}
aline_tbl$day_28_flag <- factor(aline_tbl$day_28_flag, levels=c(0, 1))
aline_tbl$gender <- factor(aline_tbl$gender, levels=c('F', 'M'))
aline_tbl$day_icu_intime <- factor(aline_tbl$day_icu_intime)
aline_tbl$hour_icu_intime <- factor(aline_tbl$hour_icu_intime)
aline_tbl$icu_hour_flag <- factor(aline_tbl$icu_hour_flag, levels=c(0, 1))
aline_tbl$sedative_flag <- factor(aline_tbl$sedative_flag, levels=c(0, 1))
aline_tbl$fentanyl_flag <- factor(aline_tbl$fentanyl_flag, levels=c(0, 1))
aline_tbl$midazolam_flag <- factor(aline_tbl$midazolam_flag, levels=c(0, 1))
aline_tbl$propofol_flag <- factor(aline_tbl$propofol_flag, levels=c(0, 1))
aline_tbl$chf_flag <- factor(aline_tbl$chf_flag, levels=c(0, 1))
aline_tbl$afib_flag <- factor(aline_tbl$afib_flag, levels=c(0, 1))
aline_tbl$renal_flag <- factor(aline_tbl$renal_flag, levels=c(0, 1))
aline_tbl$liver_flag <- factor(aline_tbl$liver_flag, levels=c(0, 1))
aline_tbl$copd_flag <- factor(aline_tbl$copd_flag, levels=c(0, 1))
aline_tbl$cad_flag <- factor(aline_tbl$cad_flag, levels=c(0, 1))
aline_tbl$stroke_flag <- factor(aline_tbl$stroke_flag, levels=c(0, 1))
aline_tbl$malignancy_flag <- factor(aline_tbl$malignancy_flag, levels=c(0, 1))
aline_tbl$respfail_flag <- factor(aline_tbl$respfail_flag, levels=c(0, 1))
aline_tbl$ards_flag <- factor(aline_tbl$ards_flag, levels=c(0, 1))
aline_tbl$pneumonia_flag <- factor(aline_tbl$pneumonia_flag, levels=c(0, 1))
aline_tbl$service_surg <- factor(aline_tbl$service_unit == 'SURG', levels=c(FALSE, TRUE))

aline_tbl <- aline_tbl %>% select(aline_flag, 
                  age, gender, weight_first, sofa_first, service_surg, 
                  day_icu_intime, hour_icu_intime, 
                  chf_flag, afib_flag, renal_flag, 
                  liver_flag, copd_flag, cad_flag, stroke_flag, 
                  malignancy_flag, respfail_flag, 
                  map_first, hr_first, temp_first, spo2_first, 
                  bun_first, chloride_first, creatinine_first, 
                  hgb_first, platelet_first, 
                  potassium_first, sodium_first, tco2_first, wbc_first,
                  day_28_flag)
```


__We make the assumption that adjustment for $L$ ensures conditional exchangeability.__
  
We will use inverse probability weights (IPW) to construct a __marginal structural model.__

The weights will be defined as follows:
$$W^a = \frac{1}{f(A|L)}$$

The denominator will be estimated by way of the following logistic regression
model:
$$\sigma(Pr[A=1|L]) = \beta_0 + \beta_LL$$

```{r}
propensity_glm <- glm(aline_flag ~ ., family = binomial(link = 'logit'), 
                      data = aline_tbl %>% select(-day_28_flag))
prA1L <- predict(propensity_glm, aline_tbl, type = 'response')
prA0L <- 1 - prA1L
```

The following modeling assumptions apply to this model:  
1) With respect to the continuous variables in $L$, the association between 
these variables is assumed linear with the log odds of the probability of being treated.  
2) The contribution of all of the covariates in $L$ to the probability of
recieving an IAC is additive on the log odds scale.   

We then assign the appropriate weight to each subject.
```{r}
aline_tbl <- aline_tbl %>% 
  mutate(wa = ifelse(aline_flag == 1, 1/prA1L, 1/prA0L))
```

We want to estimate the parameters of the following marginal structural model:
$$\sigma(Pr[Y^a]) = \theta_0 + \theta_1A$$

We estimate the parameters of the marginal structural model by fitting the 
following logistic regression model weighted by our IPW:
$$\sigma(Pr[Y|A]) = \theta_0 + \theta_1A$$

```{r warning=FALSE, message=FALSE}
msm_glm <- glm(day_28_flag ~ aline_flag, weights = wa, 
               family = binomial(link = 'logit'), data = aline_tbl)
```

This final model is non-parametric, and thus no _modeling_ assumptions were made.  

Before we examine our result we also want to construct 95% confidence intervals.
However, we have violated a fundamental assumptions of generalized linear models:
independence of the samples; we have done this via weighting. Because some samples
count for more than 1 and others for less than 1, we will have correlation which
will artificially shrink our confidence intervals. We will therefore construct a
robust confidence interval. For this take we will use the sandwich estimator
to estimate the variance robustly and then construct the condience intervals.

```{r}
se <- sqrt(diag(vcovHC(msm_glm, type = 'HC0')))
upper_ci <- coef(msm_glm) + qnorm((1 - 0.95)/2, lower.tail = F)*se
lower_ci <- coef(msm_glm) - qnorm((1 - 0.95)/2, lower.tail = F)*se
```

With these calculated, we output the final result.
```{r}
result <- data.frame(beta_1 = round(exp(coef(msm_glm)[2]), 4), round(exp(lower_ci[2]), 4), 
           round(exp(upper_ci[2]), 4))
result %>%
  knitr::kable(col.names = c('Estimate', 'Lower 95% CI', 'Upper 95% CI'), 
               row.names = F)
```

__Interpretation:__ The the odds of death at 28 days had everyone in the study 
recieved an IAC was `r result[1,1]` times the odds of death had everyone in the 
study not recieved an IAC was equal to  mmHg during the study period. If we 
were to repeat this study 100 times we would expect that 95 of the produced 
estimates would fall within the range defined by [`r result[1,2]`, `r result[1,3]`].








